---
title: "Signature Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Signature Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  fig.path = "./images/", 
  message = TRUE,
  warning = TRUE,
  error = TRUE,
  collapse = TRUE,
  comment = "#>"
)

# Load R packages
# For data cleaning, extraction and manipulation
library(tidyverse)

# For loading and installing packages
library(devtools)

# Load SigRepo package
devtools::load_all()

# Load OmicSignature package
library("OmicSignature")

```

This tutorial will go over how to interact with the core functions of the **SigRepo** package. 
Please navigate to the [README](https://github.com/montilab/SigRepo) on our GitHub repo to setup an account and create your connection handler in R-Studio.

## Installation

- Using `devtools` package

```
# Load devtools package
library(devtools)

# Install SigRepo
devtools::install_github(repo = 'montilab/SigRepo')

# Install OmicSignature
devtools::install_github(repo = 'montilab/OmicSignature')

# Load tidyverse package
library(tidyverse)

# Load SigRepo package
library(SigRepo)

# Load OmicSignature package
library(OmicSignature)
```

## Load packages 

```
# Load tidyverse package
library(tidyverse)

# Load SigRepo package
library(SigRepo)

# Load OmicSignature package
library(OmicSignature)
```

## Create connection handler

```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
# Get the test connection handler
conn_handler <- SigRepo::test_conn_handler
```

```
# Create a connection handler
conn_handler <- SigRepo::newConnHandler(
  dbname = "sigrepo", 
  host = "sigrepo.org", 
  port = 3306, 
  user = <your_username>, 
  password = <your_password>
)
``` 

## Load Signatures

Here, we provide two signature objects that comes with the package for demonstrations:

1. LLFS_Aging_Gene_2023 
2. Myc_reduce_mice_liver_24m

```{r, warning=FALSE, echo=FALSE, message=FALSE}
# Read in the signature objects
utils::data("LLFS_Aging_Gene_2023", package = "SigRepo")
utils::data("Myc_reduce_mice_liver_24m", package = "SigRepo")
```

## Upload a signature

The `SigRepo::addSignature()` function allows users to upload a signature to the database. 

__IMPORTANT NOTE:__ 

- User `MUST HAVE` an `editor` or `admin` account to use this function. 
- A signature `MUST BE` an R6 object obtained from `OmicSignature::OmicSignature()`

## **Example 1**: Upload `LLFS_Aging_Gene_2023` signature

```{r Example 1}
## Show the OmicSignature summary
print(LLFS_Aging_Gene_2023)

## Add the signature to the repository
SigRepo::addSignature(
  conn_handler = conn_handler, 
  omic_signature = LLFS_Aging_Gene_2023
)
```

## **Example 2**: Upload `Myc_reduce_mice_liver_24m` signature
In this example, we show what happens when the OmicSignature contains `feature_name`'s that are not part of "dictionary" for the corresponding omics layer.

```{r Example 2}
## Show the OmicSignature summary
print(Myc_reduce_mice_liver_24m)

missing_features <- SigRepo::addSignature(
  conn_handler = conn_handler, 
  omic_signature = Myc_reduce_mice_liver_24m,
  return_missing_features = TRUE       # Whether to return a list of missing features during upload.
)
```

## **Example 3**: Create an omic signature using **OmicSignature** package and upload to the database 

```{r Example 3}
# Create an OmicSignature metadata
metadata <- OmicSignature::createMetadata(
  # required attributes:
  signature_name = "Myc_reduce_mice_liver_24m_readme",
  organism = "Mus musculus",
  direction_type = "bi-directional",
  assay_type = "transcriptomics",
  phenotype = "Myc_reduce",

  # optional and recommended:
  covariates = "none",
  description = "mice Myc haploinsufficient (Myc(+/-))",
  platform = "transcriptomics by array",
  sample_type = "liver", # use BRENDA ontology

  # optional cut-off attributes.
  # specifying them can facilitate the extraction of signatures.
  logfc_cutoff = NULL,
  p_value_cutoff = NULL,
  adj_p_cutoff = 0.05,
  score_cutoff = 5,

  # other optional built-in attributes:
  keywords = "Myc, KO, longevity",
  cutoff_description = NULL,
  author = NULL,
  PMID = "25619689",
  year = 2015,

  # example of customized attributes:
  others = base::list("animal_strain" = "C57BL/6")
)

# Create difexp object
difexp <- base::readRDS(base::file.path(base::system.file("extdata", package = "OmicSignature"), "difmatrix_Myc_mice_liver_24m.rds")) 
base::colnames(difexp) <- OmicSignature::replaceDifexpCol(base::colnames(difexp))

# Rename ensembl with feature name and add group label to difexp
difexp <- difexp |>  
  dplyr::rename(feature_name = ensembl) |> 
  dplyr::mutate(group_label = base::as.factor(base::ifelse(.data$score > 0, "MYC Reduce", "WT")))

# Create signature object
signature <- difexp |>
  dplyr::filter(base::abs(.data$score) > metadata$score_cutoff & .data$adj_p < metadata$adj_p_cutoff) |>
  dplyr::select(c("probe_id", "feature_name", "score")) |>
  dplyr::mutate(group_label = base::as.factor(base::ifelse(.data$score > 0, "MYC Reduce", "WT")))

# Create signature object 
omic_signature <- OmicSignature::OmicSignature$new(
  metadata = metadata,
  signature = signature,
  difexp = difexp
)

# Add signature to database
SigRepo::addSignature(
  conn_handler = conn_handler,        # A handler contains user credentials to establish connection to a remote database
  omic_signature = omic_signature,    # An R6 object obtained from OmicSignature::OmicSignature()
  visibility = FALSE,                 # Whether to make signature public or private. Default is FALSE.
  return_signature_id = FALSE,        # Whether to return the uploaded signature id. Default is FALSE.
  verbose = TRUE                      # Whether to print diagnostic messages. Default is TRUE.
)
```

# Search for a list of signatures

The `SigRepo::searchSignature()` function allows users to search for all or a specific set of signatures that are available in the database. This function is a bit more nuanced than it might seem. When you search for the signature, you will be given the metadata of the signature.This does not mean you have access to download the Signature. The best way to think of this is like a library. You can browse the catalog of books, but some you might need certain permissions to access.

## Example 1: Search for all signatures. 
 
```{r}
# Get all signatures
signature_tbl <- SigRepo::searchSignature(conn_handler = conn_handler)

# WARNINGS: THE SIGNATURE TABLE CAN BE LARGE, SO ONLY THE FIRST SIX OBSERVATIONS ARE SHOWN.
if(base::nrow(signature_tbl) > 0){
  knitr::kable(
    utils::head(signature_tbl), 
    row.names = FALSE
  )
}
```

## Example 2: Search for a specific signature, e.g., `signature_name = "LLFS_Aging_Gene_2023"`

```{r} 
signature_tbl <- SigRepo::searchSignature(
  conn_handler = conn_handler, 
  signature_name = "LLFS_Aging_Gene_2023"
)

if(base::nrow(signature_tbl) > 0){
  knitr::kable(
    signature_tbl, 
    row.names = FALSE
  )
}
```

# Retrieve a list of omic signatures

The `SigRepo::getSignature()` function allows users to retrieve a list of omic signature objects that they are `PUBLICLY` available in the database. 

__IMPORTANT NOTE:__ 

- Users can `ONLY RETRIEVE` a list of signatures that are publicly available in the database including their own uploaded signatures. 
- If a signature is `PRIVATE` and belongs to other user in the database, users will need to be given an `editor` permission from its owner in order to access, retrieve, and edit their signatures.

## Example 1: Retrieve all signatures that are publicly available or owned by the user in the database 

```
# WARNING: THE SIGNATURE LIST CAN BE LARGE TO DOWNLOAD (NOT RUN)
#signature_list <- SigRepo::getSignature(conn_handler = conn_handler)
```

## Example 2: Retrieve a specific signature that is publicly available or owned by the user in the database, e.g., `signature_name = "LLFS_Aging_Gene_2023"`

```{r}
LLFS_oms <- SigRepo::getSignature(
  conn_handler = conn_handler, 
  signature_name = "LLFS_Aging_Gene_2023"
)
```

# Delete a signature

The `SigRepo::deleteSignature()` function allows users to delete a signature from the database. 

__IMPORTANT NOTE:__ 

- Users `MUST HAVE` an `editor` or `admin` account to use this function. 
- Users can `ONLY DELETE` their own uploaded signatures or were given an `editor` permission from its owner to access, retrieve, and edit their signatures.
- Users can `ONLY DELETE` a signature one at a time.

**For example:** You want to remove `signature_name = "LLFS_Aging_Gene_2023"` from the database.

```{r}
# Let's search for signature_name = "LLFS_Aging_Gene_2023" in the database
signature_tbl <- SigRepo::searchSignature(
  conn_handler = conn_handler, 
  signature_name = "LLFS_Aging_Gene_2023"
)

# If the signature exists, remove it from the database
if(base::nrow(signature_tbl) > 0){
  SigRepo::deleteSignature(
    conn_handler = conn_handler, 
    signature_id = signature_tbl$signature_id  
  )
}
```

# Update a signature 

The `SigRepo::updateSignature()` function allows users to update a specific signature in the SigRepo database. 

__IMPORTANT NOTE:__ 

- Users `MUST HAVE` an `editor` or `admin` account to use this function. 
- Users can `ONLY UPDATE` their own uploaded signatures. Only a signature's owner or an `admin` can update it; being given `editor` permission to someone else's signature is not enough.
- If an update fails part-way, the original signature is restored, including the users it is shared with and the collections it belongs to.
- Users can `ONLY UPDATE` a signature one at a time.

**For example:** If the `platform` information in the previous uploaded signature, `"Myc_reduce_mice_liver_24m_readme"`, is incorrect, and you wish to update the `platform` information with a correct value, e.g., `platform = "transcriptomics by single-cell RNA-seq"`. You can use the `SigRepo::updateSignature()` function as follows:

```{r}
# Revise the metadata object with new platform = "transcriptomics by single-cell RNA-seq"
metadata_revised <- OmicSignature::createMetadata(
  # required attributes:
  signature_name = "Myc_reduce_mice_liver_24m_readme",
  organism = "Mus musculus",
  direction_type = "bi-directional",
  assay_type = "transcriptomics",
  phenotype = "Myc_reduce",

  # optional and recommended:
  covariates = "none",
  description = "mice Myc haploinsufficient (Myc(+/-))",
  platform = "transcriptomics by single-cell RNA-seq",
  sample_type = "liver", # use BRENDA ontology

  # optional cut-off attributes.
  # specifying them can facilitate the extraction of signatures.
  logfc_cutoff = NULL,
  p_value_cutoff = NULL,
  adj_p_cutoff = 0.05,
  score_cutoff = 5,

  # other optional built-in attributes:
  keywords = "Myc, KO, longevity",
  cutoff_description = NULL,
  author = NULL,
  PMID = "25619689",
  year = 2015,

  # example of customized attributes:
  others = base::list("animal_strain" = "C57BL/6")
)

# Create difexp object
difexp <- base::readRDS(base::file.path(base::system.file("extdata", package = "OmicSignature"), "difmatrix_Myc_mice_liver_24m.rds")) 
base::colnames(difexp) <- OmicSignature::replaceDifexpCol(base::colnames(difexp))

# Rename ensembl with feature name and add group label to difexp
difexp <- difexp |>  
  dplyr::rename(feature_name = ensembl) |> 
  dplyr::mutate(group_label = base::as.factor(base::ifelse(.data$score > 0, "MYC Reduce", "WT")))

# Create signature object
signature <- difexp |>
  dplyr::filter(base::abs(.data$score) > metadata_revised$score_cutoff & .data$adj_p < metadata_revised$adj_p_cutoff) |>
  dplyr::select(c("probe_id", "feature_name", "score")) |>
  dplyr::mutate(group_label = base::as.factor(base::ifelse(.data$score > 0, "MYC Reduce", "WT")))

# Create signature object 
updated_omic_signature <- OmicSignature::OmicSignature$new(
  metadata = metadata_revised,
  signature = signature,
  difexp = difexp
)
```

```{r}
# Now, let's search for Myc_reduce_mice_liver_24m_readme in the database
# in which we would like to revise the value of platform to "transcriptomics by single-cell RNA-seq"
signature_tbl <- SigRepo::searchSignature(
  conn_handler = conn_handler, 
  signature_name = metadata_revised$signature_name
)

# If signature exists, update the signature with the revised omic_signature object
if(base::nrow(signature_tbl) > 0){
  SigRepo::updateSignature(
    conn_handler = conn_handler, 
    signature_id = signature_tbl$signature_id, 
    omic_signature = updated_omic_signature
  )
}
```

Let's look up `signature_name = "Myc_reduce_mice_liver_24m_readme"` and see if the value of `platform` has been changed.

```{r}
signature_tbl <- SigRepo::searchSignature(
  conn_handler = conn_handler, 
  signature_name = "Myc_reduce_mice_liver_24m_readme"
)

if(base::nrow(signature_tbl) > 0){
  knitr::kable(
    signature_tbl,
    row.names = FALSE
  )
}
```

Finally, remove `signature_name = "Myc_reduce_mice_liver_24m_readme"` from the database

```{r}
# Let's search for signature_name = "Myc_reduce_mice_liver_24m_readme" in the database
signature_tbl <- SigRepo::searchSignature(
  conn_handler = conn_handler, 
  signature_name = "Myc_reduce_mice_liver_24m_readme"
)

# If the signature exists, remove it from the database
if(base::nrow(signature_tbl) > 0){
  SigRepo::deleteSignature(
    conn_handler = conn_handler, 
    signature_id = signature_tbl$signature_id
  )
}
```
